Tails or Types? A Critical Systematic Review of Taxometric Studies in Neurodevelopmental Disorders

### 2025-08-09 Paper/Paper.qmd
computationally reproducible paper redone

---
title: "Tails or Types? A Critical Systematic Review of Taxometric Studies in Neurodevelopmental Disorders"
shorttitle: "Taxometric Review Neurodevelopmental Disorders"
authors:
  - name: Enrico Toffalini
    corresponding: true
    orcid: 0000-0002-1404-5133
    email: enrico.toffalini@unipd.it
    affiliation: Department of General Psychology, University of Padua, Italy
  - name: Margherita Calderan
    corresponding: false
    orcid: 0009-0005-5668-5162
    email: margherita.calderan@unipd.it
    affiliation: Department of General Psychology, University of Padua, Italy
  - name: Riccardo Pagan
    corresponding: false
    email: riccardo.pagan.3@studenti.unipd.it
    affiliation: Department of Social and Developmental Psychology, University of Padua, Italy
  - name: Valentina Tobia
    corresponding: false
    email: tobia.valentina@unisr.it
    affiliation: University Vita-Salute San Raffaele, Milan, Italy
author-note:
  disclosures:
    study-registration: 'The review was preregistered at https://osf.io/65y9g/?view_only=91e97c6b00dc444abb83fefcf67cbac0'
    financial-support: 'Financial support: Project 2022KBW99S "Tails or types? Testing the dimensional hypothesis in neurodevelopmental disorders", funded by Next Generation EU, Mission 4, Componente 1, CUP C53D23004210006'
    gratitude: "We are grateful to Athanasia Vasileiadou who first pointed us to the topic, and to some of the reviewed articles."
    data-sharing: https://osf.io/ys5ad/?view_only=0e2812acfabf43fc8af01f883c7331ba
    related-report: null
    conflict-of-interest: null
    authorship-agreements: null
abstract: "Taxometric analysis is explicitly designed to test whether the latent structures underlying clinical psychological constructs are categorical (taxonic) or dimensional. Despite its relevance to ongoing debates about the nature of neurodevelopmental disorders (NDDs), its application in this field has been limited. We conducted a preregistered systematic review to critically evaluate the literature, with a focus on methodological threats such as artificial admixture. All published taxometric studies on autism, ADHD, learning disorders, and language impairment were identified and assessed. Studies on ADHD and language impairment generally supported dimensional structures, whereas most studies on autism and NDD subtypes predominantly reported taxonic findings. However, artificial admixture was frequently observed and undermines the credibility of many taxonic conclusions. To illustrate this risk, we conducted a Monte Carlo simulation showing how admixture can spuriously generate pseudotaxonic results. We urge caution in interpreting the outcomes, and advocate for more high-quality taxometric studies in this area."
keywords: [Neurodevelopmental disorders, Taxometric analysis, Autism, ADHD, Learning disorders]
bibliography: references.bib
format:
  apaquarto-docx: default
  apaquarto-html: default
  apaquarto-typst: default
  apaquarto-pdf:
    documentmode: man
---

@@

In recent years, research on neurodevelopmental disorders (NDDs) has undergone significant reconsideration. Traditional diagnostic categories defined in the DSM-5 [@dsm5] have increasingly been questioned in favor of a dimensional and transdiagnostic perspective [e.g., @astle2022trandiagnostic; @michelini2024world; @happe2020looking; @happe2021both; @sonuga2020kuhnian; @caviola2024editorial]. Notably, the DSM-5 itself has incorporated some dimensional features (for example by defining a broad Autism Spectrum Disorder \[ASD\] in place of distinct diagnoses). Several factors have motivated this shift: the marked heterogeneity within diagnostic groups, the weak correspondence between data-driven clusters and DSM-based categories, the frequent comorbidity and phenotypic correlations across conditions, and the view that individuals with NDDs may represent the tails of developmental continua, quantitatively rather than qualitatively different from the so-called neurotypical population [@michelini2024world; @happe2021both; @mammarella2021nocore].

Determining whether NDDs represent categories, rather than just tails of continuous distributions spanning the general population, is important for several reasons, including questions of social and personal identity for affected individuals, implications for diagnostic framing, comorbidity, and intervention choices in clinical contexts, and consequences for the organization of classrooms, educational practices, and even the design and power analysis of research studies [@caviola2024editorial]. More fundamentally, this question is relevant for our understanding of the ontological status of psychological phenomena.

A dimensional approach aligns well with long-standing principles in the study of individual differences in psychology, and has long characterized research on intelligence and personality [@caviola2024editorial]. It also seems well-suited for NDDs, given a growing body of evidence supporting polygenicity and multifactorial etiology [@plomin2005generalist; @kovas2006generalist; @pennington2012multiple; @astle2020core; @caviola2024editorial; @demontis2023adhdgenome]. Most behavioral traits are understood to result not from one or a few causes, but from the additive effects of many small influences [@chabris2015fourth]. According to the central limit theorem [@fischer2011central], when many individual causes (e.g., discrete genetic variants) contribute additively, the resulting trait tends to approximate a continuous, Gaussian distribution. Nonetheless, a principled categorical view still resists in neurodevelopmental research, for example concerning ASD [e.g., @happe2021both; @chown2021bit; @frazier2023asd], subtypes within NDDs [e.g., @burgess2018taxodyslexia] or the proposed emergence of data-driven taxa [@astle2022trandiagnostic].

Taxometric analysis is a set of quantitative methods originally introduced by Paul Meehl and colleagues and refined over decades [@meehl1995taxometric], explicitly aimed at establishing whether a clinical disorder is characterized by an underlying categorical or dimensional structure. It has been widely applied in psychopathology, and typically suggested that most mental health conditions (e.g., depression, anxiety) are better conceptualized as dimensional [@haslam2012review; @haslam2020metaanalysis]. In brief, taxometric procedures, such as MAMBAC, MAXCOV, MAXEIG, and L-Mode examine covariation patterns among observed indicators to adjudicate between competing latent categorical vs dimensional models [@ruscio2007taxometric; @ruscio2004checklist; @cole2004taxometric]. These methods now feature simulation-based tools like the Comparison Curve Fit Index (CCFI), which provide quantitative measures to facilitate inference [@ruscio2018using].

Unfortunately, the use of taxometric analysis has remained limited in neurodevelopmental research, with most existing studies focusing on adult psychopathology [@beauchaine2003taxometrics; @haslam2020metaanalysis]. In their debate on whether ASD is best conceptualized as a category or a continuum, @happe2021both and @chown2021bit do not even mention such methods. In their influential review, @astle2022trandiagnostic briefly acknowledge taxometry, but predominantly emphasize studies using cluster analysis. However, the limitations of clustering methods for assessing taxonicity have been discussed for decades: clustering algorithms tend to impose (rather than test for) categorical structures, they provide weak tests of taxonicity, and often fail to recover known taxa in simulations [@beauchaine2003taxometrics; @ruscio2004boundaryissues]. Recent simulations using psychologically realistic data (featuring both some skewness and intercorrelated indicators) suggest that commonly used clustering methods are prone to detecting pseudoclusters while presenting low power to identify true taxa [@toffalini2022entia; @toffalini2024clusters]. Indeed, @astle2022trandiagnostic interpret clustering results as evidence that data-driven groups do not align with DSM-5 categories, rather than as compelling tests of latent structural form.

Taxometric analysis is not without limitations, some of which are shared with cluster analysis. Commonly cited threats to validity of conclusions include poor indicator validity, nuisance covariance, low base rate, and artificial admixture [@cole2004taxometric; @haslam2020metaanalysis; @ruscio2004checklist; @ruscio2018using].

Poor indicator validity implies that indicators present little mean difference between individuals with and without the target condition (i.e., the putative taxon and complement). Ideally, Cohen’s *d* for each indicator should substantially exceed 1 (*d* $\gg$ 1). Nuisance covariance refers to strong residual correlations among indicators after accounting for group membership, indicating redundancy. Low base rate means that too few individuals in the sample belong to the taxon, limiting power (samples with *n* \< 30 for the taxonic group are usually deemed insufficient). This is particularly relevant when studying rare conditions. Artificial admixture occurs when participants are combined from different recruitment sources or selection criteria, often as an attempt to compensate for a low base rate. For instance, clinical cases may be recruited separately from controls and merged into a single sample. In general, poor indicator validity and low base rate reduce power to detect true taxa, while high nuisance covariance and artificial admixture increase the risk of detecting pseudotaxa by mimicking a categorical structure.

@@

The primary aim of this systematic review is to critically examine the use of taxometric methods in studies of NDDs within the published literature. We focused on the most commonly investigated NDD categories: Autism Spectrum Disorder (ASD), Attention-Deficit/Hyperactivity Disorder (ADHD), Specific Learning Disorders (SLDs), and the closely related condition of Language Disorder/Impairment. (The latter was not included in the preregistered protocol and constitutes a minor deviation.) We excluded Intellectual Disability, as its etiology is often clearly identifiable and taxonic/syndromic by definition in many cases (e.g., Down syndrome), or linked to specific medical risk factors, while residual non-syndromic cases are typically regarded as reflecting the lower tail of general mental ability and thus inherently dimensional.

A crucial goal was to assess the methodological quality and credibility of taxometric findings. We specifically evaluated potential sources of bias, such as artificial admixture, and assessed whether studies provided sufficient methodological information to evaluate other known threats to validity, including low indicator validity, high nuisance covariance, and skewed indicator distributions. (Strong skewness can create the illusion of a categorical structure, and it might be relevant when assessing rare or extreme behaviors.)

Following the review, we report a Monte Carlo simulation study designed to empirically illustrate the extent to which artificial admixture can lead to the spurious detection of categorical latent structures in otherwise dimensional data.

# Systematic Review of Taxometric Analysis on NDDs

## Methods

The systematic review was conducted in accordance with the Preferred Reporting items for Systematic Reviews and Meta-Analysis (PRISMA) 2020 guidelines. A review protocol was preregistered on OSF registries at the following link: https://osf.io/65y9g/?view_only=91e97c6b00dc444abb83fefcf67cbac0 The methods are detailed below.

@@

A systematic search for relevant peer-reviewed articles was conducted in April 2025 using four electronic databases: PsycINFO, PubMed, Scopus, and Web of Science. No time limits were applied, although records predating 1990 were not expected, as taxometric methods were not yet widely adopted at that time. The following search query—reflecting the methods and target populations of interest—was applied to titles, abstracts, and keywords:

@@

In databases where the wildcard `*` was not supported (e.g., PubMed), it was replaced with full terms (e.g., `"disorder"`, `"disability"` instead of `"dis*"`; `"mathematical"`, `"mathematics"`, `"math"` instead of `"math*"`).

@@

Additionally, the full search query was launched on Google Scholar, and the first 10 pages of results were screened manually. References cited in, and articles citing, @haslam2020metaanalysis were also screened via Google Scholar. Finally, an exploratory search using OpenAI’s GPT-4o was conducted, using the preregistration protocol as contextual input. All promising records identified through these additional methods and not retrieved via database searches were listed under the *"Identification of studies via other methods"* section of the PRISMA flowchart.

@@

Inclusion criteria were as follows:\
1. The study applied at least one taxometric procedure;\
2. The target population consisted of individuals with NDDs or related traits;\
3. The study drew a conclusion about the latent structure (categorical vs. dimensional), or reported sufficient information (e.g., CCFIs) to infer such a conclusion.

Exclusion criteria were:\
1. The article was theoretical, methodological, or a simulation study without empirical data;\
2. The study did not report any behavioral data (in practice, no screened study was excluded for this criterion);\
3. The study was a review and/or meta-analysis;\
4. The study employed latent structure methods other than taxometric analysis (e.g., only clustering), with no taxometric procedures reported.

In addition, following the preregistered protocol, the search was limited to published, peer-reviewed journal articles. This decision reflected the aim of providing a critical review of the existing empirical literature. Only one potentially relevant non-journal record was identified and, while excluded from the systematic review, is discussed separately in the *Limitation* section.

@@

Two independent human coders (MC and RP) screened and coded all articles. No AI tools conducted any screening or coding, although coders occasionally used them to help clarify specific article content. All discrepancies were resolved through discussion and were later reviewed by the PI (ET).

Since a single article could report multiple (sub)samples and/or multiple taxometric analyses on the same sample, the full coded dataset followed a long-form data structure, with some articles appearing across multiple rows. Coded variables were defined in the data dictionary included in the preregistration protocol, with a few additional fields added to capture coder notes and agreement metrics.

Inter-coder agreement was fair to good across all phases. In the title and abstract screening phase, Cohen’s *κ* = 0.79 (34 out of 38 independent decisions matched). In the full-text eligibility screening, Cohen’s *κ* = 0.52 (15 out of 19). For the final data extraction, 84% of rows (71 out of 85) were extracted by both coders. Agreement on critical variables was high: for the "authors' taxonic conclusion" agreement was 94%; for "artificial admixture", it was 95% at the sample or subsample level.

## Results

A total of 110 records were identified through database searches. After removing 67 duplicates and excluding 5 additional records (dissertations or book chapters, only one of which appeared to include primary taxometric analyses but could not be retrieved), 38 unique articles were screened at the title and abstract level.

Nineteen records were excluded at this stage. Full-text assessment was then conducted on the remaining 19 articles, of which 7 were excluded for not meeting the inclusion criteria: 4 did not apply taxometric methods, and 3 did not target neurodevelopmental disorders.

An additional 4 records were identified through citation tracking and online browsing. Of these, 2 were excluded (1 was a methodological article, and 1 did not employ taxometric analysis).

In total, 14 studies met inclusion criteria and were included in the review. The PRISMA flowchart in @fig-prisma summarizes the study selection process.

@@

A summary of the included studies, simplified to one row per article, is presented in @tbl-summary. A full reporting of all coded data for each individual taxometric analysis is available in the Supplemental Online Material. The total number of taxometric analyses was `r nrow(df)`, ranging from `r min(table(df$ID_article))` to `r max(table(df$ID_article))` analyses per article. The targeted disorders were ADHD (6 articles), ASD (6 articles), dyslexia (1 article), and Specific Language Impairment (SLI; 2 articles), with one article addressing both ADHD and ASD. Most articles (10 out of 14) examined whether a given disorder constituted a taxon, while the remaining 4 focused on whether putative subtypes within a disorder exhibited taxonic structure.

@@

Taxonic conclusions were mixed. All five articles targeting ADHD as a disorder concluded in favor of a dimensional structure, as did both articles focusing on SLI. In contrast, all four articles examining ASD as a disorder leaned toward, or explicitly supported, a categorical (taxonic) structure. The remaining four articles investigated subtypes of disorders, and all leaned toward taxonic conclusions—either strongly or predominantly [@stevens2018functional on ADHD subtypes; @munson2008asd on ASD subtypes; @obrien2012dyslexia on dyslexia subtypes], or partially [@ingram2008defining, which supported ASD subtypes as taxa depending on the set of indicators used]. In total, 8 out of 14 articles reached a taxonic conclusion in at least some of their taxometric analyses. Specifically, `r sum(df$Taxonic_conclusion)` out of `r nrow(df)` total analyses—representing `r round(mean(df$Taxonic_conclusion) * 100, 1)`%—yielded taxonic results. These were primarily associated with ASD: `r sum(grepl("ASD", df$Target_disorder) & df$Taxonic_conclusion == 1)` out of `r sum(grepl("ASD", df$Target_disorder))` analyses involving ASD (i.e., `r round(100 * sum(grepl("ASD", df$Target_disorder) & df$Taxonic_conclusion == 1) / sum(grepl("ASD", df$Target_disorder)), 1)`%) supported a categorical structure.

Artificial admixture was evident or strongly suspected in at least part of the analyses in 8 out of 14 articles, involving a total of `r sum(df$Artificial_admixture)` out of `r nrow(df)` analyses. Notably, articles that incurred artificial admixture tended to align with those that reached taxonic conclusions: 6 of the 8 articles with suspected admixture also reported taxonic findings. This association was even more pronounced at the level of individual analyses. A taxonic conclusion was reached in `r sum(df$Taxonic_conclusion[df$Artificial_admixture == 1])` out of `r sum(df$Artificial_admixture == 1)` analyses (`r round(100 * mean(df$Taxonic_conclusion[df$Artificial_admixture == 1]), 1)`%) where admixture was evident or suspected. In contrast, only `r sum(df$Taxonic_conclusion[df$Artificial_admixture == 0])` out of `r sum(df$Artificial_admixture == 0)` analyses (`r round(100 * mean(df$Taxonic_conclusion[df$Artificial_admixture == 0]), 1)`%) reached a taxonic conclusion when admixture was likely or certainly absent.

Remaining methodological issues concerned missing information about key descriptive statistics and validity indices. In particular, **nuisance covariance**—that is, correlations among indicators within diagnostic groups or within the putative taxon and complement—could be coded for only 2 out of 14 articles. Where reported, it was small to negligible in @james2016latent, but exceeded 0.40 for many indicators in @frazier2023asd; notably, both studies reached a taxonic conclusion. **Skewness** of indicator score distributions was reported in only 5 articles. When available, skewness values were often near or beyond ±1 for at least one indicator in roughly one-third of analyses. However, there was no clear association between skewness and whether a taxonic conclusion was reached. By contrast, **indicator validity** was more consistently reported, and could be coded in all but 3 articles. Where available, validity (typically expressed as Cohen’s *d*) was almost always large or very large (*d* $\gg$ 1), consistent with best-practice recommendations for taxometric analysis.

@@

Artificial admixture represents a powerful threat to validity that is common but also easily avoidable by design. Different types of admixture can be incurred. @ruscio2004checklist describe three cases. First, when clinical and nonclinical participants are separately recruited and merged into a single sample. Second, when a sample is artificially split into subsamples based on different selection criteria, and these are then used to test different taxometric questions [cf. "subtractive" compound sampling, @haslam2020metaanalysis]. Third, when researchers trim observations from the putative complement (individuals without a diagnosis) to inflate the base rate of the taxon. Each of these procedures risks imposing a categorical structure onto data that might be dimensional in nature. Importantly, admixture poses a threat not only to taxometric analysis, but also to clustering methods, for the exact same reason.

>>>>

---
title: "Tails or Types? A Critical Systematic Review of Taxometric Studies in Neurodevelopmental Disorders"
shorttitle: "Taxometric Review Neurodevelopmental Disorders"
authors:
  - name: Enrico Toffalini
    corresponding: true
    orcid: 0000-0002-1404-5133
    email: enrico.toffalini@unipd.it
    affiliation: Department of General Psychology, University of Padua, Italy
  - name: Margherita Calderan
    corresponding: false
    orcid: 0009-0005-5668-5162
    email: margherita.calderan@unipd.it
    affiliation: Department of General Psychology, University of Padua, Italy
  - name: Riccardo Pagan
    corresponding: false
    email: riccardo.pagan.3@studenti.unipd.it
    affiliation: Department of Social and Developmental Psychology, University of Padua, Italy
  - name: Valentina Tobia
    corresponding: false
    email: tobia.valentina@unisr.it
    affiliation: University Vita-Salute San Raffaele, Milan, Italy
  - name: Edmund Sonuga-Barke
    corresponding: false
    email: edmund.sonuga-barke@kcl.ac.uk
    affiliation: King’s College London, UK
author-note:
  disclosures:
    study-registration: 'Review preregistered protocol: https://osf.io/65y9g/?view_only=91e97c6b00dc444abb83fefcf67cbac0'
    financial-support: 'Financial support: Project 2022KBW99S "Tails or types? Testing the dimensional hypothesis in neurodevelopmental disorders", funded by Next Generation EU, Mission 4, Componente 1, CUP C53D23004210006'
    gratitude: "We are grateful to Athanasia Vasileiadou who first pointed us to the topic, and to some of the reviewed articles."
    data-sharing: https://osf.io/ys5ad/?view_only=0e2812acfabf43fc8af01f883c7331ba
    related-report: null
    conflict-of-interest: null
    authorship-agreements: null
abstract: "**Background.** Taxometric analysis is specifically designed to test whether the latent structures underlying clinical conditions are categorical (taxonic) or dimensional in nature. Despite its relevance to ongoing debates about the nature of neurodevelopmental conditions, its application to this field has been limited. **Methods.** We conducted a preregistered systematic review of published taxometric studies on autism, ADHD, learning disorders, and language impairment, with a focus on methodological threats such as artificial admixture (i.e., the practice of combining subsamples recruited via alternative methods, such as clinical and control groups, into a single dataset). A Monte Carlo simulation was run to illustrate how admixture can yield false taxonic results from dimensional data. **Results.** Fourteen articles met the inclusion criteria. Studies of ADHD and language impairment largely supported dimensional models, whereas most studies of autism and conditions’ subtypes supported categorical models. However, admixture was common in studies reaching taxonic conclusions, especially for autism. In simulations, admixture was consistently associated with taxonic findings despite the true structure being dimensional. **Conclusions.** Evidence for categorical structure, especially in autism and disorders’ subtypes, is likely undermined by methodological limitations, particularly admixture. We advocate for more high-quality taxometric studies in this area."
keywords: [Neurodevelopmental disorders, Taxometric analysis, Autism, ADHD, Learning disorders]
bibliography: references.bib
format:
  apaquarto-docx: default
  apaquarto-html: default
  apaquarto-typst: default
  apaquarto-pdf:
    documentmode: man
---

@@

In recent years, there has been a reconsideration of the conceptual foundations of neurodevelopmental conditions (NDCs). The notion that diagnostic entities defined in the DSM-5 reflect discrete categories with identifiable non-arbitrary ‘natural’ boundaries,   separating disorder from non-disorder, and one disorder from another [@dsm5] has increasingly been questioned in favor of dimensional and transdiagnostic perspectives [e.g., @astle2022trandiagnostic; @caviola2024editorial; @happe2020looking; @happe2021both; @mammarella2021nocore; @michelini2024world; @posner2020adhd; @sonuga2020kuhnian]. In fact, the issue has been discussed for a long time. @sonuga-barke1998categoricalmodels questioned whether a medical model that frames childhood conditions as categorical disorders intrinsically linked to dysfunction was appropriate and useful. While diagnostic manuals such as the DSM are presented as atheoretical, they in practice adopt a "neo-Kraepelinian" vision of psychopathology, treating disorders as natural disease entities [see also @coghill2012categories].

A dimensional perspective reinterprets NDCs as extreme variations along continuous multivariate traits that are at least quasi-normally distributed in the general population. Accordingly, individuals with NDCs are not categorically distinct, and differ quantitatively rather than qualitatively from the neurotypical population. From this perspective diagnostic thresholds are essentially arbitrary, determined not by natural discontinuities but by pragmatic considerations such as clinical need [@arildskov2022howmuch; @haslam2006latentstructure].

Determining whether NDCs represent tails of continuous distributions or discrete categories with boundaries grounded in biology or clinical reality, is important for our understanding of the ontological status of neurodevelopmental phenomena and for assessing the adequacy of the medical dysfunction paradigm. If NDCs are dimensional in nature, then the diagnostic thresholds used to categorize disorders are, by definition, arbitrary, reframing diagnostic categories as socially constructed groupings that include individuals at the margins of trait continua. The implications potentially affect how diagnosed individuals understand their social identity, how clinicians make decisions, how classrooms and support systems are organized to accommodate neurodivergence, and how research is designed and interpreted [@caviola2024editorial].

Support for a dimensional account of NDCs arises from several converging lines of evidence. First, individuals sharing the same diagnostic label often show substantial heterogeneity in symptom profiles, cognitive functioning, and developmental trajectories [e.g., @astle2022trandiagnostic]. Second, the boundaries between diagnostic categories are frequently blurred, with high comorbidity [@coghill2012categories; @michelini2024world]. Third, dimensional models are consistent with the growing evidence supporting multifactorial and polygenic etiology in NDCs [@astle2020core; @caviola2024editorial; @kovas2006generalist; @pennington2012multiple; @plomin2005generalist]. According to the central limit theorem [@fischer2011central], when many individual causes (e.g., genetic variants) contribute additively, the resulting trait tends to approximate a continuous, Gaussian distribution. Despite this body of evidence, categorical thinking still dominates parts of the neurodevelopmental literature, for example concerning the categorical nature of ASD [e.g., @chown2021bit; @frazier2023asd; @happe2021both], subtypes within NDCs [e.g., @burgess2018taxodyslexia] or the proposed interpretation of emerging data-driven taxa [@astle2022trandiagnostic].

While much of the evidence for dimensionality is circumstantial, taxometric analyses, originally introduced by Paul Meehl and colleagues [@meehl1995taxometric], were developed to provide a direct empirical test of whether latent clinical construct are categorical (taxonic) or dimensional in nature. Taxometric methods evaluate the structure of covariance among observed variables to determine whether there exists a latent boundary that divides individuals into qualitatively distinct groups. Conceptually, if such a boundary exists, it should alter and disrupt the patterns of relationships between indicators. Otherwise, the patterns are smooth and continuous across the entire range of observed values, supporting a dimensional model. Frequently used taxometric procedures include MAMBAC (Mean Above Minus Below A Cut), MAXCOV (Maximum Covariance), MAXEIG (Maximum Eigenvalue), and L-Mode (Latent Mode Factor Analysis) [@cole2004taxometric; @ruscio2007taxometric; @ruscio2004boundaryissues]. Recently, simulation-based tools such as the Comparison Curve Fit Index (CCFI) have been introduced to provide objective, quantitative indices of fit between observed data and simulated taxonic versus dimensional comparison curves, facilitating inference [@ruscio2018using].

Taxometric analyses have been widely applied in adult psychopathology, generally supporting dimensional structures for most adult psychiatric conditions [@haslam2012review; @haslam2020metaanalysis]. However, their use in neurodevelopmental research has remained limited, though the importance of taxometric methods as explicit tests for determining whether childhood disorders are categorical or dimensional has been repeatedly emphasized [@sonuga-barke1998categoricalmodels; @beauchaine2003taxometrics; @coghill2012categories]. In their debate on whether ASD reflects a category or a continuum, @happe2021both and @chown2021bit do not mention such methods. In their review, @astle2022trandiagnostic briefly acknowledge taxometry, but predominantly emphasize studies using cluster analysis (a family of data-driven methods aimed at grouping individuals based on similarity across multiple variables). However, the limitations of clustering methods for assessing taxonicity have been discussed for decades: clustering algorithms tend to impose (rather than test for) categorical structures, they provide weak tests of taxonicity, and often fail to recover known taxa in simulations [@beauchaine2003taxometrics; @ruscio2004boundaryissues]. Recent simulations using psychometrically realistic data suggest that commonly used clustering methods are prone to inflate the number of detected clusters [@toffalini2022entia; @toffalini2024clusters]. Indeed, @astle2022trandiagnostic interpreted clustering results as evidence that data-driven groups fail to align with DSM-5 categories, rather than as strong tests of taxonicity.

Like any method, taxometric analysis has its limitations. Commonly cited sources of bias in taxometric conclusions include poor indicator validity, nuisance covariance, low base rate, and artificial admixture [@cole2004taxometric; @haslam2020metaanalysis; @ruscio2018using; @ruscio2004boundaryissues; @ruscio2004checklist]. Poor indicator validity implies that indicators present little mean difference between individuals with and without the target condition. Ideally, Cohen’s *d* for each indicator should substantially exceed 1. Nuisance covariance refers to strong residual correlations among indicators after accounting for group membership, indicating redundant information. Low base rate means that too few individuals in the sample belong to the taxon, limiting power to detect it (*n \>* 30 for the taxonic group is usually considered necessary), which is problematic when studying rare conditions. Artificial admixture occurs when participants are combined from different recruitment sources or selection criteria, often as an attempt to compensate for a low base rate. @ruscio2004checklist describe three admixture-related sampling practices that may lead to pseudotaxonic results. First, when clinical and nonclinical participants are separately recruited and merged into a single sample. Second, when a sample is artificially split into subsamples based on different selection criteria, and these are then used to test different taxometric questions [cf. "subtractive" compound sampling, @haslam2020metaanalysis]. Third, when researchers trim observations from the putative complement (individuals without a diagnosis) to inflate the base rate of the taxon. All these scenarios tend to artificially impose categorical structures onto data that might be strictly dimensional.

@@

We aimed to critically review the use of taxometric methods in published studies of NDCs. We focused on the most commonly investigated categories: Autism Spectrum Disorder (ASD), Attention-Deficit/Hyperactivity Disorder (ADHD), Learning Disorders, and the closely related condition of Language Impairment/Disorder (the latter constitutes a minor deviation from the preregistration protocol). We excluded Intellectual Disability, whose etiology is often identifiable and taxonic/syndromic by definition (e.g., Down syndrome), or linked to known medical risk factors.

A crucial goal was to assess the methodological quality and credibility of taxometric findings. We specifically evaluated potential sources of bias, such as artificial admixture, and assessed whether studies provided sufficient methodological information to evaluate other known threats to validity, including low indicator validity, high nuisance covariance, and indicators skewness [extreme skewness can also create the illusion of categorical structures, @ruscio2004boundaryissues; @ruscio2004checklist, and it might be relevant when assessing rare or extreme behaviors].

Finally, we conduct a Monte Carlo simulation study to empirically illustrate the extent to which artificial admixture can lead to the risk to reaching taxonic conclusions in otherwise dimensional data. We simulate a dimensional scenario where clinical and nonclinical participants are separately recruited and merged into a single sample; taxometric (and model-based cluster) analyses are then performed on a set of observed indicators reflecting the latent clinical dimension.

# Methods

The systematic review was conducted in accordance with the *Preferred Reporting items for Systematic Reviews and Meta-Analysis* (PRISMA) 2020 guidelines. A review protocol was preregistered on OSF registries at the following link: https://osf.io/65y9g/?view_only=91e97c6b00dc444abb83fefcf67cbac0

@@

A systematic search for relevant peer-reviewed articles was conducted in April 2025 using four electronic databases: PsycINFO, PubMed, Scopus, and Web of Science. No time limits were applied. The following search query, reflecting the methods and populations of interest, was applied to titles, abstracts, keywords:

@@

In PubMed, where the wildcard `*` was not supported, it was replaced with full terms (e.g., `"disorder"`, `"disability"` instead of `"dis*"`; `"mathematical"`, `"mathematics"`, `"math"` instead of `"math*"`).

@@

Additionally, the full search query was launched on Google Scholar, and the first 10 pages of results were screened manually. References cited in, and articles citing, @haslam2020metaanalysis were also screened via Google Scholar. Finally, an exploratory search using OpenAI’s GPT-4o “Deep Research” tool was conducted, providing the preregistration protocol as context. All additional promising records identified through these methods were listed under the *“Identification of studies via other methods”* section of the flowchart.

@@

Inclusion criteria were:

1.  The study applied at least one taxometric procedure (MAMBAC, MAXCOV, MAXEIG, L-Mode, MAXSLOPE; other procedures would be considered if explicitly described as taxometric);
2.  The target population consisted of individuals with NDCs/traits;
3.  The study drew a conclusion about the latent structure (categorical vs. dimensional) or reported sufficient information (e.g., CCFIs) to infer a conclusion.

Exclusion criteria were:

1.  The article was theoretical, methodological, simulation study without empirical data;
2.  The study did not analyze behavioral data;
3.  The study was a review and/or meta-analysis.

The search was limited to published, peer-reviewed journal articles, as we aimed to provide a critical review of the existing high-quality literature. Only one potentially relevant non-journal record was identified and, while excluded from the systematic review, is commented separately in the Discussion.

@@

Two independent human coders (MC and RP) screened and coded all articles. All discrepancies were resolved through discussion and were later reviewed by the PI (ET).

Since a single article could report multiple (sub)samples and/or taxometric analyses, the full coded dataset followed a long-form data structure, with articles appearing across multiple rows. Coded variables were defined in the data dictionary included in the preregistration protocol, with a few additional fields added to capture coder notes and agreement metrics.

Inter-coder agreement was fair to good across all phases. In the title and abstract screening phase, Cohen’s *κ*=0.79 (34 out of 38 independent decisions matched). In the full-text eligibility screening, Cohen’s *κ*=0.52 (15 out of 19). For the final data extraction, 84% of rows (71 out of 85) were extracted by both coders. Agreement on critical variables was high: for the “authors’ taxonic conclusion” agreement was 94%; for “artificial admixture”, it was 95% at the (sub)sample level.

# Results

A total of 110 records were identified through database searches. After removing 67 duplicates and excluding 5 additional records (dissertations or book chapters, only one of which appeared to include primary taxometric analyses but could not be retrieved), 38 articles were screened for title/abstract. Nineteen records were excluded at this stage. Full-text assessment was then conducted on the remaining 19 articles, of which 7 were excluded for not meeting the inclusion criteria. An additional 4 records were identified through citation tracking and online browsing, of which 2 were included in the final review. The PRISMA flowchart in @fig-prisma summarizes the study selection process.

@@

A summary of the included studies, simplified to one row per article, is presented in @tbl-summary. A full reporting of all coded data for each individual taxometric analysis is available in the Supplemental Online Material. The total number of taxometric analyses was `r nrow(df)`, ranging from `r min(table(df$ID_article))` to `r max(table(df$ID_article))` per article. Targeted disorders were ADHD (6 articles), ASD (6 articles), dyslexia (1 article), and Specific Language Impairment (SLI; 2 articles), with one article addressing both ADHD and ASD. Most articles (10 out of 14) examined whether a given disorder constituted a taxon, while the remaining 4 focused on whether putative subtypes within a disorder showed taxonic structure.

@@

Taxonic conclusions were mixed. All five articles targeting ADHD as a disorder concluded in favor of a dimensional structure, as did both articles focusing on SLI. In contrast, all four articles examining ASD as a disorder leaned toward, or explicitly supported, a categorical (taxonic) structure. The remaining four articles investigated subtypes of disorders, and all favored taxonic conclusions, either predominantly [@stevens2018functional on ADHD subtypes; @munson2008asd on ASD subtypes; @obrien2012dyslexia on dyslexia subtypes], or partially [@ingram2008defining supported taxonic ASD subtypes depending on the set of indicators used]. In total, 8 out of 14 articles reached a taxonic conclusion in at least some of their taxometric analyses. Specifically, `r sum(df$Taxonic_conclusion)` out of `r nrow(df)` total analyses—representing `r round(mean(df$Taxonic_conclusion) * 100, 1)`%—yielded taxonic results. These were primarily associated with ASD: `r sum(grepl("ASD", df$Target_disorder) & df$Taxonic_conclusion == 1)` out of `r sum(grepl("ASD", df$Target_disorder))` analyses involving ASD (i.e., `r round(100 * sum(grepl("ASD", df$Target_disorder) & df$Taxonic_conclusion == 1) / sum(grepl("ASD", df$Target_disorder)), 1)`%) supported a categorical structure.

Artificial admixture was evident or strongly suspected in at least part of the analyses in 8 out of 14 articles (including all those targeting ASD) and involving a total of `r sum(df$Artificial_admixture)` out of `r nrow(df)` analyses. Articles that incurred artificial admixture tended to align with those that reached taxonic conclusions: 6 of the 8 articles with suspected admixture also reported taxonic findings. This association was even more pronounced at the level of individual analyses. A taxonic conclusion was reached in `r sum(df$Taxonic_conclusion[df$Artificial_admixture == 1])` out of `r sum(df$Artificial_admixture == 1)` analyses (`r round(100 * mean(df$Taxonic_conclusion[df$Artificial_admixture == 1]), 1)`%) where admixture was evident or suspected. In contrast, only `r sum(df$Taxonic_conclusion[df$Artificial_admixture == 0])` out of `r sum(df$Artificial_admixture == 0)` analyses (`r round(100 * mean(df$Taxonic_conclusion[df$Artificial_admixture == 0]), 1)`%) reached a taxonic conclusion when admixture was likely or certainly

Other methodological issues concerned missing information about key descriptive statistics and validity indices. Nuisance covariance could be coded for only 2 out of 14 articles. Where reported, it was small to negligible in @james2016latent , but exceeded 0.40 for many indicators in @frazier2023asd . Both studies reached a taxonic conclusion. Skewness of indicator score distributions was reported in only 5 articles, where skewness values were often near or beyond ±1 for at least one indicator in about one-third of analyses. However, there was no clear association between skewness and whether a taxonic conclusion was reached. By contrast, indicator validity was more consistently reported, and could be coded in all but 3 articles. Where available, it was almost always large or very large (Cohen’s *d* ≫ 1), consistent with best-practice recommendations for taxometric analysis.

@@

Artificial admixture represents a common and powerful threat to validity, but also one that is easily avoidable by design. In our review, it presented a marked risk of bias for studies supporting the taxonic nature of autism. To illustrate the risks associated with artificial admixture, we simulate a dimensional scenario where clinical and nonclinical participants are separately recruited and merged into a single sample
